### 2020-05-01 probabilisticProgrammingMartingales.tex
Proof of Theorem 1: ongoing

\newcommand{\Real}{\mathbb{R}}
\newcommand\expect[1]{\mathbb{E}[#1]}
\newcommand\set[1]{\{#1\}}

@@

%\begin{proof}
Given a closed term $M$ and a ranking function $f$ for it, define random variables on the probability space $S$ (where $s$ is a random variable) by
\begin{align*}
(M_n,s_n) & = \red^n(M,s) \\
y_0 & = 0 \\
y_{n+1} & = \min \{ k \mid k>y_n, M_k \text{ a value or of the form } E[\tY N] \}\\
M'_n & = M_{y_n} \\
X_n & = f(M'_n)
\end{align*}
and define a filtration $\mathcal{F}_n = \sigma(M_k, k \leq n)$ (i.e. all the samples used up to step $n$).

@@

Next we clearly have 
\(
f(M_{n+1}) = %\sum_{i=1}^3 f_{i, n+1}(M).
f_{1, n+1}(M) + f_{2, n+1}(M) + f_{3, n+1}(M),
\)
$\mu$-almost-surely.
(Unless otherwise stated, all equations and inequations between random variables are assumed to hold only $\mu$-a.s.)
Therefore 
\begin{align}
& \expect{f(M_{n+1}) \mid \mathcal{F}_n}
\nonumber \\ 
& = \expect{f_{1, n+1}(M) + f_{2, n+1}(M) + f_{3, n+1}(M)  \mid \mathcal{F}_n} 
\nonumber \\
& = \expect{f_{1, n+1}(M) \mid \mathcal{F}_n} + \expect{f_{2, n+1}(M) \mid \mathcal{F}_n} + \expect{f_{3, n+1}(M) \mid \mathcal{F}_n} 
\label{eqn:linear cond ex} \\
& = f_{1, n+1}(M) + f_{2, n+1}(M) + f_{3, n+1}(M)
\label{eqn:inde cond exp} \\
%& = f_{1, n+1}(M) + \int_I f() + f_{3, n+1}(M) \label{eqn:inde cond exp} \\
& \leq f(M_{n})[T_1] + f(M_{n})[T_2] + f(M_{n})[T_3]
\label{eqn:ts} \\
& = f(M_n). 
\label{eqn:partition} 
\end{align}
\Cref{eqn:linear cond ex} follows from the linearity of conditional expectation. 
\Cref{eqn:inde cond exp} is justified because $f_{i, n+1}(M)$ is $\mathcal{F}_n$-measurable (\Cref{lem:inde}).
Inequation~(\ref{eqn:ts}) follows from $f_{i, n+1}(M) \leq f(M_n)[T_i]$ for each $i$. 
Here we use the notation: given function $g : A \to \Real$ and $U \subseteq A$, $g[U] : A \to \Real$ denotes the function
\[
g[U](x) :=
\begin{cases}
g(x) & \textrm{$x \in U$}\\
0 & \textrm{o/w}
\end{cases}
\]
and write $T_i = \{s \in S \mid \textrm{$M_n(s)$ is type-$i$}\}$.
To see $f_{2, n+1}(M) \leq f(M_n)[T_2]$,
take $s \in T_2$. 
Then $f_{2, n+1}(M)(s) = f(E[\underline{a}])$ and $M_n = E[\tsample]$, for some $a \in [0, 1]$ and evaluation context $E$.
Hence 
\[
f_{2, n+1}(M)(s) \leq \int_I f(E[\underline{r}]) \, \mu_{leb}(\textrm{d} r)
\leq f(E[\tsample]) = f(M_n)[T_2](s).
\]
Finally, \Cref{eqn:partition} holds because $\{T_1, T_2, T_3, T_4\}$ is a partition of $S$.
}

>>>>

\newcommand{\Real}{\mathbb{R}}
\newcommand\expect[1]{\mathbb{E}[#1]}
\newcommand\set[1]{\{#1\}}
\newcommand\dif{\mathrm{d}}

@@

%\begin{proof}

\lo{We also need to show $f(M_n)$ is integrable.}

Given a closed term $M$ and a ranking function $f$ for it, define random variables on the probability space $S$ (where $s$ is a random variable) by
\begin{align*}
(M_n,s_n) & = \red^n(M,s) \\
y_0 & = 0 \\
y_{n+1} & = \min \{ k \mid k>y_n, M_k \text{ a value or of the form } E[\tY N] \}\\
M'_n & = M_{y_n} \\
X_n & = f(M'_n)
\end{align*}
and define a filtration $\mathcal{F}_n = \sigma(M_k, k \leq n)$ (i.e. all the samples used up to step $n$).

@@

Next we clearly have 
\(
f(M_{n+1}) = %\sum_{i=1}^3 f_{i, n+1}(M).
f_{1, n+1}(M) + f_{2, n+1}(M) + f_{3, n+1}(M),
\)
$\mu$-almost-surely.
(Unless otherwise stated, all equations and inequations between random variables are assumed to hold only $\mu$-a.s.)
Therefore 
\begin{align}
& \expect{f(M_{n+1}) \mid \mathcal{F}_n}
\nonumber \\ 
& = \expect{f_{1, n+1}(M) + f_{2, n+1}(M) + f_{3, n+1}(M)  \mid \mathcal{F}_n} 
\nonumber \\
& = \expect{f_{1, n+1}(M) \mid \mathcal{F}_n} + \expect{f_{2, n+1}(M) \mid \mathcal{F}_n} + \expect{f_{3, n+1}(M) \mid \mathcal{F}_n} 
\label{eqn:linear cond ex} \\
& = f_{1, n+1}(M) + f_{2, n+1}(M) + f_{3, n+1}(M)
\label{eqn:inde cond exp} \\
%& = f_{1, n+1}(M) + \int_I f() + f_{3, n+1}(M) \label{eqn:inde cond exp} \\
& \leq f(M_{n})[T_1] + f(M_{n})[T_2] + f(M_{n})[T_3]
\label{eqn:ts} \\
& = f(M_n). 
\label{eqn:partition} 
\end{align}
\Cref{eqn:linear cond ex} follows from the linearity of conditional expectation. 
\Cref{eqn:inde cond exp} is justified because $f_{i, n+1}(M)$ is $\mathcal{F}_n$-measurable (\Cref{lem:inde}).
Inequation~(\ref{eqn:ts}) follows from $f_{i, n+1}(M) \leq f(M_n)[T_i]$ for each $i$. 
Here we use the notation: given function $g : A \to \Real$ and $U \subseteq A$, $g[U] : A \to \Real$ denotes the function
\[
g[U](x) :=
\begin{cases}
g(x) & \textrm{$x \in U$}\\
0 & \textrm{o/w}
\end{cases}
\]
and write $T_i = \{s \in S \mid \textrm{$M_n(s)$ is type-$i$}\}$.
[** To show $\expect{f(M_{n+1}) \mid \mathcal{F}_n} \leq f(M_n)$ a.s., it suffices to show
\(\int_A \expect{f(M_{n+1}) \mid \mathcal{F}_n} \dif \mu \leq \int_A f(M_n) \dif \mu \), for all $A \in \mathcal{F}_n$. **]
% To see $f_{2, n+1}(M) \leq f(M_n)[T_2]$, take $s \in T_2$. Then $f_{2, n+1}(M)(s) = f(E[\underline{a}])$ and $M_n = E[\tsample]$, for some $a \in [0, 1]$ and evaluation context $E$. Hence 
% \[
% f_{2, n+1}(M)(s) \leq \int_I f(E[\underline{r}]) \, \mu_{leb}(\textrm{d} r)
% \leq f(E[\tsample]) = f(M_n)[T_2](s).
% \]
Finally, \Cref{eqn:partition} holds because $\{T_1, T_2, T_3, T_4\}$ is a partition of $S$.
}
